---
editor_options:
  markdown:
    wrap: 72
---

## Standing Live Detection

Christopher Tsz hin Choi NREL 03/26/2025

------------------------------------------------------------------------

### Packages

```{r}
library(sf)
library(terra)
library(corrplot)
library(caret)  # For confusionMatrix function
library(lidR)
```

### Data

```{r}

##### Vector
##################

HPCC = sf::read_sf('../data/vector/HPCC_BurnPerimeter.shp')


## Training Points
# standing.live = sf::read_sf('./data/vector/Standing_Live_Training_Data.shp')
# standing.dead = sf::read_sf('./data/vector/Standing_Dead_Training_Data.shp')
trees = sf::read_sf('../data/vector/Training_Points.shp')


grass = sf::read_sf('../data/vector/Training_Points_Grass.shp')
shadow = sf::read_sf('../data/vector/Training_Points_Shadow.shp')
bare.ground = sf::read_sf('../data/vector/Training_Points_Bare_Ground.shp')
buildings = sf::read_sf('../data/vector/Training_Points_Buildings.shp')


# training.points = rbind(standing.live, standing.dead, grass, shadow, bare.ground, buildings)
training.points = rbind(trees, grass, shadow, bare.ground, buildings)


##### Raster
###################

## Single USFS Ortho image
usfs.ortho = terra::mosaic(terra::rast('../data/raster/Ortho_HermitsPeak2022_3610554_sw_3.tif'),
                           terra::rast('../data/raster/Ortho_HermitsPeak2022_3610554_sw_4.tif'))

## LiDAR
CHM = terra::rast('../data/raster/CHM_Crop.tif') |>
  terra::crop(usfs.ortho)



# ## Temp 
# 
# ## Single USFS Ortho image
# usfs.ortho = terra::mosaic(terra::rast('C:/Github/HPCC-Tree-Survival-Project/output/Ortho_HermitsPeak2022_3610554_sw_3.tif'),
# terra::rast('C:/Github/HPCC-Tree-Survival-Project/output/Ortho_HermitsPeak2022_3610554_sw_4.tif'))
# 
# CHM = terra::rast('2023_LiDAR/Canopy_Height_Model_2023_Raster/HPCC_2023_chm.tif') |>
#   terra::crop(usfs.ortho)
# 
# terra::writeRaster(CHM, 'C:/Github/HPCC-Tree-Survival-Project/output/CHM_crop.tif')
# terra::writeRaster(usfs.ortho, 'C:/Github/HPCC-Tree-Survival-Project/output/usfs.ortho.tif')



```

Plot data
```{r}
# plot(NDVI)
# plot(HPCC$geometry, add = T)


plot(CHM)
plot(HPCC$geometry, add = T)
```



## Identify Tree Tops
```{r}
# Generate kernel and smooth chm
kernel <- matrix(1, 3, 3)
schm <- terra::focal(x = CHM, w = kernel, fun = median, na.rm = TRUE)
plot(schm)
```

```{r}
# Detect trees
ttops <- lidR::locate_trees(las = schm, algorithm = lmf(ws = 2.5))
ttops
```

## Raster Processing

### Separate bands and create indices

```{r}
RED = usfs.ortho$Ortho_HermitsPeak2022_3610554_sw_3_1
names(RED) <- "RED" ## Rename band to recognizable name

GREEN = usfs.ortho$Ortho_HermitsPeak2022_3610554_sw_3_2
names(GREEN) <- "GREEN"

BLUE = usfs.ortho$Ortho_HermitsPeak2022_3610554_sw_3_3
names(BLUE) <- "BLUE"

NIR = usfs.ortho$Ortho_HermitsPeak2022_3610554_sw_3_4
names(NIR) <- "NIR"

# Normalized Difference Vegetation Index
NDVI = (NIR - RED)/(NIR + RED)
names(NDVI) <- "NDVI"

# Enhanced Vegetation Index
EVI = (2.5*(NIR - RED)) / ((NIR+(6*RED)-(7.5*BLUE)) + 1)
names(EVI) <- "EVI"

# Soil adjusted vegetation index
SAVI = ((NIR - RED) / (NIR + RED + 0.5)) * 1.5
names(SAVI) <- "SAVI"


names(CHM) <- "CHM"

dat = training.points |>
  dplyr::mutate(RED = (terra::extract(RED, training.points)[,2]),
                GREEN = (terra::extract(GREEN, training.points)[,2]),
                BLUE = (terra::extract(BLUE, training.points)[,2]),
                NIR = (terra::extract(NIR, training.points)[,2]),
                NDVI = (terra::extract(NDVI, training.points)[,2]),
                EVI = (terra::extract(EVI, training.points)[,2]),
                SAVI = (terra::extract(SAVI, training.points)[,2]),
                CHM = (terra::extract(CHM, training.points)[,2]))
```

### Data exploration (Testing for correlation)

```{r}
corrplot::corrplot(cor(as.data.frame(dat) |> dplyr::select(RED, 
                                               GREEN,
                                               BLUE,
                                               NIR,
                                               NDVI,
                                               EVI,
                                               SAVI,
                                               CHM)),
                    method="number",shade.col=NA, tl.col="black", tl.srt=45)
```

## Modeling

Modeling using logistic regression because we are primarily trying to
determine whether a tree is live or dead.

```{r}
logit.dat = training.points |>
  dplyr::mutate(TreeCondBinary = as.factor(ifelse(TreeCond == "Live", 1, 0))) |>
  cbind(dat)

set.seed(123)  # Ensures reproducibility
split.index <- sample(1:nrow(dat), size = 0.7 * nrow(dat))  # Randomly select 70% of rows

train.data <- logit.dat[split.index, ]  # 70% training data
test.data <- logit.dat[-split.index, ]  # 30% testing data

## Multiple Logistic Regression
## Using this model because we are primarily trying to determine weather a tree is live or dead



model1 <- glm(as.factor(TreeCond) ~ RED + GREEN + BLUE + NIR + NDVI + EVI + SAVI + CHM, data = train.data, family = "binomial")
model2 <- glm(as.factor(TreeCond) ~ RED + GREEN + BLUE + EVI + CHM, data = train.data, family = "binomial")
model3 <- glm(as.factor(TreeCond) ~ RED + NIR + NDVI + EVI + CHM, data = train.data, family = "binomial")
model4 <- glm(as.factor(TreeCond) ~ RED +  NIR + EVI + CHM, data = train.data, family = "binomial")



# summary(model1)
# summary(model2)
# summary(model3)
summary(model4)

```

Model Validation

```{r}

test.data = as.data.frame(test.data)

# Predict probabilities on the test data
test.data$PredictedProbs <- predict(model4, newdata = test.data, type = "response")

# Convert probabilities to binary predictions (threshold = 0.5)
test.data$PredictedClass <- ifelse(test.data$PredictedProbs > 0.5, 1, 0)
```

Confusion Matrix

```{r}
# Convert to factors for confusion matrix
test.data$TreeCond <- as.factor(test.data$TreeCond)
test.data$PredictedClass <- as.factor(test.data$PredictedClass)

# Generate confusion matrix
conf.matrix <- caret::confusionMatrix(test.data$PredictedClass, test.data$TreeCondBinary)
print(conf.matrix)
```

## Apply Model to Region

### Create Raster Stack and convert to dataframe

```{r}
# raster.stack = c(terra::mask(RED, HPCC))

raster.stack = c(RED |> terra::mask(HPCC),
                 EVI |> terra::mask(HPCC),
                 NIR |> terra::mask(HPCC),
                 CHM |> terra::resample(RED) |> terra::mask(HPCC))


raster.df = as.data.frame(raster.stack, na.rm = T)

```

### Predict using raster dataframe

```{r}
## Predict probabilities
raster.df$PredictedProbs <- predict(model4, newdata = raster.df, type = 'response')

## Convert probabilities to binary class
raster.df$PredictedProbsBinary <- ifelse(raster.df$PredictedProbs > 0.5, 1, 0)
```

### Apply to raster

```{r}
predicted.raster <- RED ## Use RED reaster as template
terra::values(predicted.raster) <- raster.df$PredictedProbs ## Assign prediction

# writeRaster(predicted.raster, "Ouputs/Tree_Condition_Prediction_Test.tif", foramt = 'GTiff', overwrite = T)

```


Plot raster prediction
```{r}
plot(predicted.raster)
plot(HPCC$geometry, add = T)

```



### Model on Tree tops dataset

Extract raster values using ttops
```{r}
ttops = ttops |>
  dplyr::mutate(RED = (terra::extract(RED, ttops)[,2]),
                GREEN = (terra::extract(GREEN, ttops)[,2]),
                BLUE = (terra::extract(BLUE, ttops)[,2]),
                NIR = (terra::extract(NIR, ttops)[,2]),
                NDVI = (terra::extract(NDVI, ttops)[,2]),
                EVI = (terra::extract(EVI, ttops)[,2]),
                SAVI = (terra::extract(SAVI, ttops)[,2]),
                CHM = (terra::extract(CHM, ttops)[,2]))

```

```{r}
## Predict probabilities
ttops$predicted <- predict(model4, newdata = ttops, type = 'response')
ttops$predictedBinary <- ifelse(ttops$predicted > 0.5, 1, 0)

ttops

sf::write_sf(ttops, 'C:/Github/HPCC-Tree-Survival-Project/output/Classifed_Tree_Tops.shp', layer_options = 'SHPT=POINT')
```

